Turned on!": Momentum in Round-Based First Person Shooter Esports

### 2022-07-18 presentation.qmd
rename and rerender stuff

cod_rounds <- qs::qread('cod_rounds.qs')
```

@@

::: fragment
-   Plant/Defuse in Valorant
-   Valorant professional league, updated through January 1, 2022
-   5v5
-   First to 13 round wins
-   One team plays offense in 12 consecutive cod_rounds before swapping sides
-   Must win by 2 cod_rounds
:::

@@

::: fragment
-   Total cod_rounds: **`r cod_n_rounds`**
:::

@@

## What are your chances of winning a round, given the series state?

@@

heatmap_seq_labeller <- function(.df, .label, .num) {
  sprintf(
    '%s\n(%s/%s)',
    scales::percent(.df[[.label]], accuracy = 1),
    scales::comma(.df[[.num]]),
    scales::comma(.df$n)
  )
}

@@

heatmap_text_layers <- function(
    ..., 
    .df,
    .threshold,
    .labeller,
    .f
) {
  list(
    ...,
    .f(.df = .df, .threshold = .threshold, .labeller = .labeller, .emphasize = 'low'),
    .f(.df = .df, .threshold = .threshold, .labeller = .labeller, .emphasize = 'high')
  )
}

o_round_win_text_layer <- function(
    ..., 
    .df,
    .threshold,
    .labeller, 
    .emphasize
) {
  if(.emphasize == 'low') {
    op <- `<`
    .color <- blackish_background
  } else {
    op <- `>=`
    .color <- 'white'
  }
  list(
    ...,
    heatmap_text_layer(
      .df = .df,
      .op = op,
      .label = 'win_round_prop',
      .num = 'win_round',
      .color = .color,
      .threshold = .threshold,
      .labeller = .labeller
    )
  )
}

p_cod_o_round_win_prop <- cod_o_win_prop |> 
  ggplot() +
  common_cod_heatmap_layers() +
  heatmap_tile(.fill = 'win_round_prop') +
  heatmap_text_layers(
    .df = cod_o_win_prop,
    .f = o_round_win_text_layer,
    .threshold = 0.52,
    .labeller = heatmap_seq_labeller
  ) +
  ggsci::scale_fill_material('teal') +
  labs(
    title = 'Offensive Round Win %'
  )
p_cod_o_round_win_prop
save_for_slide(p_cod_o_round_win_prop)
```

@@

## What are your chances of winning a round, given the series state?

```{r}
#| label: p_cod_o_round_win_prop_over_expected
# I have no idea why I need to add .num here... keeps erroring otherwise
heatmap_div_labeller <- function(.df, .label, .num) {
  sprintf(
    '%+.1f\n(%s)', 
    100 * .df[[.label]], 
    scales::comma(.df$n)
  )
}

@@

@@

p_cod_o_series_win_prop <- cod_o_win_prop |> 
  ggplot() +
  common_cod_heatmap_layers() +
  heatmap_tile(.fill = 'win_series_prop') +
  heatmap_text_layers(
    .df = cod_o_win_prop, 
    .f = o_series_win_text_layer,
    .threshold = 0.8,
    .labeller = heatmap_seq_labeller
  ) +
  ggsci::scale_fill_material('orange') +
  labs(
    title = 'Offensive Series Win %'
  )
p_cod_o_series_win_prop
save_for_slide(p_cod_o_series_win_prop)
```

@@

![](o_series_win_prop_over_expected.png){fig-align="center"}

@@

p_cod_series_outcome_prop_over_expected <- cod_long_series_outcome_prop |> 
  ggplot() +
  aes(x = record, y = outcome_idx) +
  geom_tile(
    data = cod_long_series_outcome_prop |> filter(outcome_idx > 1),
    show.legend = FALSE,
    # height = 0.8,
    aes(fill = value)
  ) +
  ggsci::scale_fill_material(
    palette = 'orange'
  ) +
  ggnewscale::new_scale_fill() +
  geom_tile(
    data = cod_long_series_outcome_prop |> filter(outcome_idx == 1),
    show.legend = FALSE,
    height = 0.8,
    aes(fill = value)
  ) +
  scale_fill_gradient2(
    low = extract_ggsci_mid_color('purple'),
    high = extract_ggsci_mid_color('green'),
    mid = 'white',
    midpoint = 0,
    limits = c(
      -cod_max_diff_series_outcome_prop, 
      +cod_max_diff_series_outcome_prop
    )
  ) +
  geom_text(
    data = cod_long_series_outcome_prop |> filter(outcome_idx == 3),
    color = blackish_background,
    size = 11 / .pt,
    aes(label = sprintf('%.1f%%\n(%s)', 100 * value, scales::comma(n)))
  ) +
  geom_text(
    data = cod_long_series_outcome_prop |> filter(outcome_idx == 2),
    color = blackish_background,
    size = 11 / .pt,
    aes(label = sprintf('%.1f%%', 100 * value))
  ) +
  geom_text(
    data = cod_long_series_outcome_prop |> filter(outcome_idx == 1) |> filter(abs(value) != max(abs(value))),
    color = blackish_background,
    size = 11 / .pt,
    aes(label = sprintf('%+.1f%%', 100 * value))
  ) +
  geom_text(
    data = cod_long_series_outcome_prop |> filter(outcome_idx == 1) |> filter(abs(value) == max(abs(value))),
    color = 'white',
    size = 11 / .pt,
    aes(label = sprintf('%+.1f%%', 100 * value))
  ) +
  scale_x_discrete(
    position = 'top',
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    labels = c('Diff.', 'Expected', 'Actual'),
    breaks = 1:3,
    expand = c(0, 0)
  ) +
  theme(
    panel.grid.major = element_blank(),
    axis.title = element_text(hjust = 0.5),
    plot.title = ggtext::element_markdown(hjust = 0.5),
    axis.text = element_text(size = 16, face = 'bold')
  ) +
  labs(
    title = 'Series outcome %',
    x = NULL,
    y = NULL
  ) +
  common_cod_labs() 
p_cod_series_outcome_prop_over_expected
save_for_slide(p_cod_series_outcome_prop_over_expected)
```

>>>>

cod_rounds <- qs::qread('cod_rounds.qs')
val_rounds <- qs::qread('valorant_rounds.qs')
```

@@

::: fragment
-   Plant/Defuse in Valorant
-   Valorant professional league, updated through January 1, 2022
-   5v5
-   First to 13 round wins
-   One team plays offense in 12 consecutive rounds before swapping sides
-   Must win by 2 cod_rounds
:::

@@

::: fragment
-   Total rounds: **`r cod_n_rounds`**
:::

@@

## Valorant data overview

```{r}
#| label: descriptive-val-snd-stats
#| eval: true
val_n_series <- val_rounds |> 
  distinct(game_id) |> 
  nrow() |> 
  scales::comma()

val_n_rounds <- val_rounds |> 
  distinct(game_id, round) |> 
  nrow() |> 
  scales::comma()

val_o_win_prop_const <- val_rounds |> 
  filter(is_offense) |> 
  count(win_round) |> 
  mutate(prop = n / sum(n)) |> 
  filter(win_round) |> 
  pull(prop)
val_o_win_pct <- scales::percent(val_o_win_prop_const, accuracy = 1)
```

::: fragment
-   Total series: **`r val_n_series`**
:::

::: fragment
-   Total rounds: **`r val_n_rounds`**
:::

::: fragment
-   Offensive win %: **`r val_o_win_pct`**
:::

## What are your chances of winning a round, given the series state in COD SnD?

@@

cod_heatmap_seq_labeller <- function(.df, .label, .num) {
  sprintf(
    '%s\n(%s/%s)',
    scales::percent(.df[[.label]], accuracy = 1),
    scales::comma(.df[[.num]]),
    scales::comma(.df$n)
  )
}

@@

heatmap_text_layers <- function(
    ..., 
    .df,
    .threshold,
    .labeller,
    .f
) {
  list(
    ...,
    .f(.df = .df, .threshold = .threshold, .labeller = .labeller, .emphasize = 'low'),
    .f(.df = .df, .threshold = .threshold, .labeller = .labeller, .emphasize = 'high')
  )
}
```

```{r}
#| label: p_cod_o_round_win_prop
o_round_win_text_layer <- function(
    ..., 
    .df,
    .threshold,
    .labeller,
    .label = 'win_round_prop',
    .num = 'win_round',
    .emphasize
) {
  if(.emphasize == 'low') {
    op <- `<`
    .color <- blackish_background
  } else {
    op <- `>=`
    .color <- 'white'
  }
  list(
    ...,
    heatmap_text_layer(
      .df = .df,
      .op = op,
      .label = .label,
      .num = .num,
      .color = .color,
      .threshold = .threshold,
      .labeller = .labeller
    )
  )
}

p_cod_o_round_win_prop <- cod_o_win_prop |> 
  ggplot() +
  common_cod_heatmap_layers() +
  heatmap_tile(.fill = 'win_round_prop') +
  heatmap_text_layers(
    .df = cod_o_win_prop,
    .f = o_round_win_text_layer,
    .threshold = 0.52,
    .labeller = cod_heatmap_seq_labeller
  ) +
  ggsci::scale_fill_material('teal') +
  labs(
    title = 'Offensive Round Win %'
  )
p_cod_o_round_win_prop
save_for_slide(p_cod_o_round_win_prop)
```

@@

## What are your chances of winning a round, given the series state in COD SnD?

```{r}
#| label: p_cod_o_round_win_prop_over_expected
heatmap_div_labeller <- function(.df, .label, .num) {
  sprintf(
    '%+.1f\n(%s)', 
    100 * .df[[.label]], 
    scales::comma(.df$n)
  )
}

@@

## What are your chances of winning a round, given the series state in Valorant?

```{r}
val_round_and_series_win_prop_by_side <- val_rounds |> 
  group_by(pre_cumu_w, pre_cumu_l, is_offense) |> 
  summarize(
    n = n(),
    across(c(win_round, win_series), sum)
  ) |> 
  ungroup() |> 
  mutate(
    win_round_prop = win_round / n,
    win_series_prop = win_series / n
  )

val_o_win_prop <- val_round_and_series_win_prop_by_side |> 
  filter(is_offense) |> 
  mutate(
    diff_win_round_prop = win_round_prop - !!val_o_win_prop_const
  )

# Joshua Broas, vlr.gg
common_val_labs <- function(...) {
  list(
    ...,
    labs(
      caption = 'Pro Valorant matches, 2020 - Jan 1, 2022'
    )
  )
}

common_val_heatmap_layers <- function(...) {
  list(
    ...,
    guides(
      fill = 'none'
    ),
    scale_x_continuous(
      labels = 0:12,
      breaks = seq(0.5, 12.5, by = 1),
      expand = c(0, 0)
    ),
    scale_y_continuous(
      labels = 0:12,
      breaks = seq(0.5, 12.5, by = 1),
      expand = c(0, 0),
      sec.axis = sec_axis(
        trans = I, 
        name = ' ', 
        breaks = seq(0.5, 12.5, by = 1), 
        labels = rep('', 13)
      )
    ),
    theme(
      panel.grid.major = element_blank(),
      axis.title = element_text(hjust = 0.5),
      plot.title = ggtext::element_markdown(hjust = 0.5),
      axis.text = element_text(size = 16, face = 'bold')
    ),
    common_val_labs(),
    labs(
      x = "Offensive Team's # of Pre-Round Wins",
      y = "Defensive Team's # of Pre-Round Wins"
    )
  )
}

val_heatmap_seq_labeller <- function(.df, .label, .num) {
  sprintf(
    '%s', 
    scales::percent(.df[[.label]], accuracy = 1)
  )
}

filt_val_o_win_prop <- val_o_win_prop |> 
  filter(pre_cumu_w <= 12, pre_cumu_l <= 12)

val_pre_swap_rounds <- filt_val_o_win_prop |> 
  mutate(pre_cumu_wl = pre_cumu_w + pre_cumu_l) |> 
  filter(pre_cumu_wl == 12)

val_pre_swap_round_segments <- val_pre_swap_rounds |> 
  transmute(
    pre_cumu_w,
    across(pre_cumu_l, ~.x + 1),
    i = row_number(),
    n = 2
  ) |> 
  uncount(n) |> 
  group_by(i) |> 
  mutate(rn = row_number()) |> 
  ungroup() |> 
  mutate(
    across(pre_cumu_w, ~ifelse(rn == 2, .x + 1, .x))
  ) |> 
  add_row(
    pre_cumu_w = 13,
    pre_cumu_l = 0
  )

p_val_o_round_win_prop <- filt_val_o_win_prop |> 
  ggplot() +
  common_val_heatmap_layers() +
  heatmap_tile(.fill = 'win_round_prop') +
  geom_path(
    data = val_pre_swap_round_segments,
    color = 'white',
    size = 1.5,
    aes(
      x = pre_cumu_w, 
      y = pre_cumu_l
    )
  ) +
  heatmap_text_layers(
    .df = filt_val_o_win_prop, 
    .f = o_round_win_text_layer,
    .num = NULL,
    .threshold = 0.8,
    .labeller = val_heatmap_seq_labeller
  ) +
  ggsci::scale_fill_material('teal') +
  labs(
    title = 'Offensive Round Win %'
  )
p_val_o_round_win_prop
save_for_slide(p_val_o_round_win_prop)
```

::: notes
-   Big difference compared to COD. Lower/higher percentages in the corners are probably more indicative of gaps in team skill.
-   1-0, 1-1, 3-0, 3-1 seem to b e the biggest outliers.
:::

![](val_o_round_win_prop.png){fig-align="center"}

@@

p_cod_o_series_win_prop <- cod_o_win_prop |> 
  ggplot() +
  common_cod_heatmap_layers() +
  heatmap_tile(.fill = 'win_series_prop') +
  heatmap_text_layers(
    .df = cod_o_win_prop, 
    .f = o_series_win_text_layer,
    .threshold = 0.8,
    .labeller = cod_heatmap_seq_labeller
  ) +
  ggsci::scale_fill_material('orange') +
  labs(
    title = 'Offensive Series Win %'
  )
p_cod_o_series_win_prop
save_for_slide(p_cod_o_series_win_prop)
```

@@

![](cod_o_series_win_prop_over_expected.png){fig-align="center"}

@@

p_cod_series_outcome_prop_over_expected <- cod_long_series_outcome_prop |> 
  ggplot() +
  aes(x = record, y = outcome_idx) +
  geom_tile(
    data = cod_long_series_outcome_prop |> filter(outcome_idx > 1),
    show.legend = FALSE,
    # height = 0.8,
    aes(fill = value)
  ) +
  ggsci::scale_fill_material(
    palette = 'light-blue'
  ) +
  ggnewscale::new_scale_fill() +
  geom_tile(
    data = cod_long_series_outcome_prop |> filter(outcome_idx == 1),
    show.legend = FALSE,
    height = 0.8,
    aes(fill = value)
  ) +
  scale_fill_gradient2(
    low = extract_ggsci_mid_color('purple'),
    high = extract_ggsci_mid_color('green'),
    mid = 'white',
    midpoint = 0,
    limits = c(
      -cod_max_diff_series_outcome_prop, 
      +cod_max_diff_series_outcome_prop
    )
  ) +
  geom_text(
    data = cod_long_series_outcome_prop |> filter(outcome_idx == 3),
    color = blackish_background,
    size = 11 / .pt,
    aes(label = sprintf('%.1f%%\n(%s)', 100 * value, scales::comma(n)))
  ) +
  geom_text(
    data = cod_long_series_outcome_prop |> filter(outcome_idx == 2),
    color = blackish_background,
    size = 11 / .pt,
    aes(label = sprintf('%.1f%%', 100 * value))
  ) +
  geom_text(
    data = cod_long_series_outcome_prop |> filter(outcome_idx == 1) |> filter(abs(value) != max(abs(value))),
    color = blackish_background,
    size = 11 / .pt,
    aes(label = sprintf('%+.1f%%', 100 * value))
  ) +
  geom_text(
    data = cod_long_series_outcome_prop |> filter(outcome_idx == 1) |> filter(abs(value) == max(abs(value))),
    color = 'white',
    size = 11 / .pt,
    aes(label = sprintf('%+.1f%%', 100 * value))
  ) +
  scale_x_discrete(
    position = 'top',
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    labels = c('Diff.', 'Expected', 'Actual'),
    breaks = 1:3,
    expand = c(0, 0)
  ) +
  theme(
    panel.grid.major = element_blank(),
    axis.title = element_text(hjust = 0.5),
    plot.title = ggtext::element_markdown(hjust = 0.5),
    axis.text = element_text(size = 16, face = 'bold')
  ) +
  labs(
    title = 'Series outcome %',
    x = NULL,
    y = NULL
  ) +
  common_cod_labs() 
p_cod_series_outcome_prop_over_expected
save_for_slide(p_cod_series_outcome_prop_over_expected)
```
